<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{BlandAltmanLeh Intro}
-->

---
title: "BlandAltmanLeh Intro"
author: "Bernhard Lehnert"
date: "Wednesday, August 13, 2014"
output: html_document
---



Bland-Altman-Plots
===
Bland-Altman plots are a well established method to check the agreement of
different measurement methods or the retest-reliability of a single measurement
method. They do not come included in R but can easily be produced using R.
The BlandAltmanLeh package tries to make Bland-Altman plots even more
accessable.

You can find the underlying publications by JM Bland and DG Altman on 
http://http://www-users.york.ac.uk/~mb55/meas/ba.htm form where you can also 
download for free
the 1986 publication in Lancet, which this package is based on. Written for 
medical doctors this article is easily accessible for the non-statistician.

What's the main idea?
---
Imagine, you've measured something with measurement method A and B and you want
to compare both. Let's say
```{r}
A <- c(-0.358, 0.788, 1.23, -0.338, -0.789, -0.255, 0.645, 0.506, 
0.774, -0.511, -0.517, -0.391, 0.681, -2.037, 2.019, -0.447, 
0.122, -0.412, 1.273, -2.165)
B <- c(0.121, 1.322, 1.929, -0.339, -0.515, -0.029, 1.322, 0.951, 
0.799, -0.306, -0.158, 0.144, 1.132, -0.675, 2.534, -0.398, 0.537, 
0.173, 1.508, -1.955)
```

Your first attempt to inspect these data may be a scatter plot like
```{r}
plot(A, B)
abline(0,1)
```

Bland and Altman propose a different approach, where the x axis is the mean of
the two measurements and the y axis is the difference between them. 
```{r}
plot((A+B)/2, A-B)
```

Now three additional lines are added for the mean of the differences and 2 
(1,96 resp.) standard deviations above and below that.
```{r}
library(BlandAltmanLeh)
bland.altman.plot(A, B)
```

ggplot2
---
Of course you might be inclined to draw that using ggplot2:
```{r}
pl <- bland.altman.plot(A, B, graph.sys = "ggplot2")
```

Which is mainly a matter of taste. As you can see, 1 out of 20 data points falls
out of the 95% confidence interval depicted by the upper and lower line. That's
just what one would expect.

Of course, these lines have an error margin and Bland and Altman 1986 describe how
to compute confidence intervals for the lines. These can also be calculated
and printed with the BlandAltmanLeh package as in:
```{r}
pl <- bland.altman.plot(A, B, graph.sys="ggplot2", conf.int=.95)

# or in base-graphics:
bland.altman.plot(A, B, conf.int=.95)
```


Sunflower-Option
---

Sometimes data have ties. Imagine your test is a questionnaire which will only
ever give scores between 0 and 10 and you are checking retest-agreement:
```{r}
A <- c(7, 8, 4, 6, 4, 5, 9, 7, 5, 8, 1, 4, 5, 7, 3, 4, 4, 9, 3, 3, 
1, 4, 5, 6, 4, 7, 4, 7, 7, 5, 4, 6, 3, 4, 6, 4, 7, 4, 6, 5)
B <- c(8, 7, 4, 6, 3, 6, 9, 8, 4, 9, 0, 5, 5, 9, 3, 5, 5, 8, 3, 3, 
1, 4, 4, 7, 4, 8, 3, 7, 7, 5, 6, 7, 3, 3, 7, 3, 6, 5, 9, 5)

bland.altman.plot(A, B)
```

Obviously there is a lot of ties in these data. There are 21 data points visible
even though there are 40 data points contained. That is why the BlandAltmanLeh
packages offers a sunflower plot as the basis of a Bland-Altman plot for data
with ties:

```{r}
bland.altman.plot(A, B, sunflower=TRUE)
```

Unfortunately, this option does not exist with ggplot2 output. However, if you
want to make a plot of your own you can still use the BlandAltmanLeh package to
compute the statistics behind the Bland-Altman plot as in this little example, 
where male and female data are to be drawn in different colors:

```{r}
A <- c(-0.358, 0.788, 1.23, -0.338, -0.789, -0.255, 0.645, 0.506, 
0.774, -0.511, -0.517, -0.391, 0.681, -2.037, 2.019, -0.447, 
0.122, -0.412, 1.273, -2.165)
B <- c(0.121, 1.322, 1.929, -0.339, -0.515, -0.029, 1.322, 0.951, 
0.799, -0.306, -0.158, 0.144, 1.132, -0.675, 2.534, -0.398, 0.537, 
0.173, 1.508, -1.955)
sex <- c( 1,1,1,1,2,2,2,1,1,1,2,2,2,2,2,1,1,2,1,2)

ba.stats <- bland.altman.stats(A, B)

plot(ba.stats$means, ba.stats$diffs, col=sex, 
     sub=paste("critical difference is",ba.stats$critical.diff),
     main="make your own graph easily", ylim=c(-1.5,1.5))
abline(h = ba.stats$lines, lty=c(2,3,2), col=c("lightblue","blue","lightblue"), 
       lwd=c(3,2,3))
```

Thus, you have the full flexibility of the R graphic systems but no need to
worry about details like missing data etc.

Yet another Bland-Altman procedure? What's in a name?
===
With Bland-Altman plots being a standard analysis, one should think, there are 
lots of packages on CRAN. Yes, there are:

* Package **PairedData** has a function plotBA based on ggplot2 (and I prefer base
plots) and no stats as return value
* Package **ResearchMethods** has a function BlandAltman which focuses on a GUI and
has no return values.
* Package **epade has** a function bland.altman.ade which appears to have no return 
values.
* Package **MethComp** has a functino BlandAltman that is deprecated and a function
ba.plot which does a lot, mainly regression and from first glance seems to be 
worthy of further investigation for serious work.
* There are **probably other packages** that I did not mention. There I did not want
to give this package a name like RBlandAltman but state, that this is just 
humble one approach by one Person (Lehnert) and therefore BlandAltmanLeh

Last but not least
===
Before using this in any serious work consider version number and perform 
plausibility checks.

Enjoy!

